Large-scale text pre-training helps with dialogue act recognition, but not without fine-tuning

### 2021-03-12 reports/IWCS2021/iwcs-short-paper.tex
edit discussion

We have two main contributions.
First we find that while standard BERT pre-training is useful,
but the model performs poorly without fine-tuning (\S\ref{sec:experiment2}).
Second, we find that further pre-training with data from the target domain shows promise for dialogue, 
but the results are mixed when pre-training with a larger corpus of dialogical data from outside the target domain (\S\ref{sec:experiment3}).

@@

DAR is the task of labeling utterances with the dialogue act they perform 
from a given set of dialogue act tags.
As with other sequence labeling tasks in NLP, some notion of context is helpful in DAR.
One of the first performant machine learning models for DAR was a Hidden Markov Model 
that used various lexical and prosodic features as input \citep{stolckeDialogueActModeling2000}.
Most successful neural approaches also model some notion of context \citep[e.g.,][]{kalchbrennerRecurrentConvolutionalNeural2013,tranHierarchicalNeuralModel2017,botheContextbasedApproachDialogue2018,bothe2018conversational,zhao2018unified}.
Following previous work, we model dialogue act sequences with an RNN over utterance representations,
which are constructed by BERT.

@@

Whether or not encoder model has undergone further in-domain pre-training, 
the choice of whether to fine-tune during task training remains.
Freezing the encoder model during task training is more efficient, 
since the gradient of the loss function need only be computed for the task-specific layers.
However, fine-tuning can lead to better performance since the encoding itself is adapted to the target task and domain.

@@

\subsection{Pre-training vs. fine-tuning} \label{sec:experiment2} % Bill
First, we analyze how pre-training affects BERT's performance as an utterance encoder.
To do so, we consider the performance of DAR models with three different utterance encoders:
\begin{itemize}
  \item \texttt{BERT-FT} -- pre-trained + DAR fine-tuning 
  \item \texttt{BERT-FZ} -- pre-trained, frozen during DAR
  \item \texttt{BERT-RI} -- randomly initialized + DAR fine-tuning 
\end{itemize}

@@

The pre-trained model is more accurate than the randomly initialized model by several percentage points on both DA corpora,
suggesting that BERT's extensive pre-training does provide some useful information for recognizing different dialogue acts (Table~\ref{tab:results}).
This performance boost is much more pronounced in the macro-averaged F1 score,
which is explained by the fact that at the tag level, pre-training has a larger impact on less frequent tags.
For a disaggregated view, see Figure 1 in the supplementary materials.

A key aspiration of transfer learning is to expose the model to phenomena that are too infrequent to learn from labeled training data alone.
One explanation for pre-trained BERT's superior performance on some infrequent dialogue acts is that it has learned to represent relevant features
that the randomly initialized model was unable to learn due to their sparsity in the labeled training set.
Indeed, a simple lexical probe supports this explanation: in utterances where the pre-trained model is correct and the randomly initialized model is not,
the rarest word is 1.9 times rarer on average than is typical of corpus as a whole.

Comparing the performance of the fine-tuned encoder to that of the frozen model, 
we see that pre-training alone is not enough.
The frozen model is heavily biased towards the most frequent tags, which explains its especially poor macro-average score (Table~\ref{tab:results}).
In SWDA, for example, the model with a frozen encoder predicts one of the two most common tags (Statement-non-opinion or Acknowledge) 86\% of the time, whereas those two
tags account for only 51\% of the ground truth tags.
The model with a fine-tuned encoder is much less biased; it predicts the two most common tags only 59\% of the time.

@@

For each of the target corpora (SWDA and AMI-DA), we compare four different pre-training conditions: 
The in-domain corpus (\texttt{ID}), consisting of the AMI pre-training corpus for the AMI-DA model and the SWBD pre-traning corpus for the SWDA model; 
the cross-domain corpus (\texttt{CC}), consisting of both the AMI and SWBD pre-training corpora; 
the OpenSubtitles corpus (\texttt{OS});
and finally the standard BERT\textsubscript{BASE} model with no additional pre-training.
As before, we experiment with both frozen and fine-tuned models at the task training stage.

We performed 10 epochs of pre-training on the in-domain models and 5 epochs of pre-training on the cross-domain models so that the total amount of training data was comparable. The OpenSubtitles models trained on much more data (see Table~\ref{tab:pretraining-corpora}) and were trained for only one epoch.

In the fine-tuned condition, additional pre-training offers a modest boost in overall accuracy and a substantial boost to the macro-average scores.
Any of the dialogue corpora appear helpful, but the cross-domain corpus generally provides the largest boost.
In the frozen condition, only the very large OpenSubtitles corpus is helpful, suggesting that when adapting BERT to dialogue, the size of the corpus is more important than its quality or fidelity to the target domain.
Still, pre-training provides nowhere near the performance improvement achieved by fine-tuning on the target task.

@@

%\subsection{Experiment 4: Cross-domain pre-training (if time permits)} \label{sec:experiment4} % Bill
%  - cross-domain pre-training
\paragraph{}
Having discussed the experimental results, the final section of the paper provides broader explanations of our findings.

\section{Discussion} % add items to discuss

This may also explain why all our best models perform better on SWDA: more of the information that interlocutors and dialogue act annotators rely on is present SWDA transcripts, whereas AMI-DA annotators receive clear instructions to pay attention to the videos \citep{GuidelinesDialogueAct2005}.
This finding is consistent with that of \citet{bavelas2008gesturing} who demonstrate that in face-to-face dialogue, visual components, such as gestures, can convey information that is independent from what is conveyed by speech.

There is abundant room for further progress in determining how other speech-related information, such as prosody and disfluencies, can be incorporated into the DAR model.
\citet{stolckeDialogueActModeling2000} showed that dialogue acts can have specific prosodic manifestations that can be used to improve dialogue act classification.
Incorporating such information is crucial if models pre-trained on large-scale text corpora are to be adapted for use in dialogue applications.

Our experiments with further pre-training were inconclusive. 
Although pre-training on data from the target domain proved useful, training on a much larger corpus of dialogue-like data (OpenSubtitles) yielded less substantial improvements in the fine-tuned condition.
\citet{baoPLATOPretrainedDialogue2019} finds that further pre-training BERT on a large-scale Reddit and Twitter corpus is helpful for response selection, but given unimpressive results with subtitles, it remains an open question how well the text chat and social media domains transfer to transcribed dialogue.

Another avenue for future research is what pre-training schemes are best for the dialogue domain.
The training scheme for RoBERTa \citep{liuRoBERTaRobustlyOptimized2019}, for example, leaves out next sentence prediction as a pre-training task because they do not find that it improves performance on question answering and natural language inference tasks.
The same may not be true of dialogue, however, where sequentiality is essential.

>>>>

We have two main contributions.
First we find that while standard BERT pre-training is useful,
the model performs poorly without fine-tuning (\S\ref{sec:experiment2}).
Second, we find that further pre-training with data from the target domain shows promise for dialogue, 
but the results are mixed when pre-training with a larger corpus of dialogical data from outside the target domain (\S\ref{sec:experiment3}).

@@

DAR is the task of labeling utterances with the dialogue act they perform 
from a given set of dialogue act tags.
As with other sequence labeling tasks in NLP, some notion of context is helpful in DAR.
One of the first performant machine learning models for DAR was a Hidden Markov Model 
that used various lexical and prosodic features as input \citep{stolckeDialogueActModeling2000}.
Most successful neural approaches also model some notion of context \citep[e.g.,][]{kalchbrennerRecurrentConvolutionalNeural2013,tranHierarchicalNeuralModel2017,botheContextbasedApproachDialogue2018,bothe2018conversational,zhao2018unified}.
%Following previous work, we model dialogue act sequences with an RNN over utterance representations,
%which are constructed by BERT.

@@

Whether or not the encoder model has undergone further in-domain pre-training, 
the choice of whether to fine-tune during task training remains.
Freezing the encoder model during task training is more efficient, 
since the gradient of the loss function need only be computed for the task-specific layers.
However, fine-tuning can lead to better performance since the encoding itself is adapted to the target task and domain.

@@

\subsection{Pre-training vs. fine-tuning} \label{sec:experiment2} % Bill
First, we analyze how pre-training affects BERT's performance as an utterance encoder.
To do so, we consider the performance of DAR models with three different utterance encoders:
\begin{itemize}\setlength\itemsep{-0.5em}
  \item \texttt{BERT-FT} -- pre-trained + DAR fine-tuning 
  \item \texttt{BERT-FZ} -- pre-trained, frozen during DAR
  \item \texttt{BERT-RI} -- random init. + DAR fine-tuning 
\end{itemize}

@@

\texttt{BERT-FT} is more accurate than \texttt{BERT-RI} by several percentage points on both DA corpora,
suggesting that BERT's extensive pre-training does provide some useful information for DAR (Table~\ref{tab:results}).
This performance boost is much more pronounced in the macro-averaged F1 score,
which is explained by the fact that at the tag level, pre-training has a larger impact on less frequent tags
(see Figure 1 in the supplementary materials).

%Comparing the performance of the fine-tuned encoder to that of the frozen model, 
%we see that pre-training alone is not enough.
The \texttt{BERT-FZ} performs very poorly compared to either \texttt{BERT-FT} or \texttt{BERT-RI}, however.
It is heavily biased towards the most frequent tags, which explains its especially poor macro-average score (Table~\ref{tab:results}).
In SWDA, for example, the model with a frozen encoder predicts one of the two most common tags (Statement-non-opinion or Acknowledge) 86\% of the time, whereas those two
tags account for only 51\% of the ground truth tags.
\texttt{BERT-FT} is much less biased; it predicts the two most common tags only 59\% of the time.

@@

For each of the target corpora (SWDA and AMI-DA), we compare four different pre-training conditions: 
The in-domain corpus (\texttt{ID}), consisting of the AMI pre-training corpus for the AMI-DA model and the SWBD pre-traning corpus for the SWDA model; 
the cross-domain corpus (\texttt{CC}), consisting of both the AMI and SWBD pre-training corpora; 
and finally the OpenSubtitles corpus (\texttt{OS}).
As before, we experiment with both frozen and fine-tuned models at the task training stage.

We performed 10 epochs of pre-training on the in-domain models and 5 epochs of pre-training on the cross-domain models so that the total amount of training data was comparable. The OpenSubtitles models were trained for only one epoch but with much more total training time.

In the fine-tuned condition, additional pre-training offers a modest boost in overall accuracy and a substantial boost to the macro-average scores, with the cross-domain corpus providing the largest boost.
In the frozen condition, only the very large OpenSubtitles corpus is helpful, suggesting that when adapting BERT to dialogue, the size of the corpus is more important than its quality or fidelity to the target domain.
Still, pre-training provides nowhere near the performance improvement achieved by fine-tuning on the target task.

@@

\section{Discussion}

A key aspiration of transfer learning is to expose the model to 
phenomena that are too infrequent to learn from labeled training data alone.
We show some evidence of that here.
Pre-trained \texttt{BERT-FT} performs better on infrequent dialogue acts than \texttt{BERT-RI}, 
suggesting it draws on the extensive pre-training to represent infrequent features of those utterances.
Indeed, a simple lexical probe supports this explanation: in utterances where the pre-trained model is correct and the randomly initialized model is not,the rarest word is 1.9 times rarer on average than is typical of corpus as a whole.

In spite of that, the representations learned through pre-training are simply not performant without
task-specific fine-tuning, suggesting that they are fundamentally lacking in information that is 
important for the dialogue context.
We should note that this is in stark contrast to many other non-dialogical semantic tasks,
where frozen BERT performs on par or \emph{better} than the fine-tuned model \citep{petersTuneNotTune2019}.

By performing additional pre-training on a large dialogue-like corpus (OpenSubtitles),
we were able to raise the performance of the frozen encoder by a small amount.
This deserves further investigation. 
\citet{baoPLATOPretrainedDialogue2019} find that further pre-training BERT on a large-scale Reddit and Twitter corpus is helpful for response selection, but given the unimpressive results with subtitles, it remains an open question how well the text chat and social media domains transfer to natural dialogue.

%This may also explain why all our best models perform better on SWDA: more of the information that interlocutors and dialogue act annotators rely on is present SWDA transcripts, whereas AMI-DA annotators receive clear instructions to pay attention to the videos \citep{GuidelinesDialogueAct2005}.
%This finding is consistent with that of \citet{bavelas2008gesturing} who demonstrate that in face-to-face dialogue, visual components, such as gestures, can convey information that is independent from what is conveyed by speech.

There is also abundant room to investigate how speech-related information, such as laughter, prosody, 
and disfluencies can be incorporated into the DAR model.
\citet{stolckeDialogueActModeling2000} showed, for example, that dialogue acts can have specific prosodic manifestations that can be used to improve dialogue act classification.
Incorporating such information is crucial if models pre-trained on large-scale text corpora are to be adapted for use in dialogue applications.

%Another avenue for future research is what pre-training schemes are best for the dialogue domain.
%The training scheme for RoBERTa \citep{liuRoBERTaRobustlyOptimized2019}, for example, leaves out next sentence prediction as a pre-training task because they do not find that it improves performance on question answering and natural language inference tasks.
%The same may not be true of dialogue, however, where sequentiality is essential.
